---
title: "Abalone age prediciton"
author: "Rohith Kumar Sajja"
date: "March 6, 2020"
output: html_notebook
---

This is an R Markdown Notebook to predict age of abalone.

The goal of this project is to predict the number of rings in the abalone and thus predict the age.


# 1. Data Description


```{r}
# Installing and loading required packages
library(e1071)
library(ggplot2)
library(dplyr)
library(knitr)
library(GGally)
library(AppliedPredictiveModeling)
library(caret)

# Loading the abalone dataset
abalone_0 <- read.csv('abalone.csv', na.strings = '?')

# Structure of the dataset
print(str(abalone_0))

# Summary statistics
summary(abalone_0)


```



Encode the Rings into 3 age buckets in a new column Age.

1. Young - less than 6 rings (<7.5 years old)

2. Adult - from 6 to 13 rings (7.5 to 14.5 years old) 

3. Old - more than 13 rings (>14.5 years old) 



```{r}

# Transforming the dataset to include age bucket
abalone <- abalone_0 %>%
  mutate(Age=case_when(
    Rings %in% 1:5 ~ "Young",
    Rings %in% 6:13 ~ "Adult",
    Rings %in% 14:30 ~ "Old"
  ))

# Converting Age into factor
abalone$Age <- as.factor(abalone$Age)

# Structure of modified data set
str(abalone)


# Summary statistics of modified data set
summary(abalone)

kable(abalone[1:10,], digits = 4,format = 'markdown')

# Number of missing values
nrow(abalone[!complete.cases(abalone),])

```

# 2. Exploratory Data Analysis

### 2.1 Pair-wise correlation
```{r}
ggpairs(abalone, aes(colour = Sex, alpha = 0.1), title="Pairs plot for abalone dataset", upper = list(continuous = wrap("cor", size = 1))) + 
  theme_grey(base_size = 10)
```

We can note that there is high correlation between a few measurements, such as Diameter and Length

### 2.2 Rings

```{r}
# Plotting the frequency of abalone over number of rings
ggplot(abalone, aes(x=Rings))+geom_bar(fill="steelblue",alpha=0.6)
```


### 2.3 Sex

```{r}
# Summary statistics
summary(abalone$Wholeweight)

# Density plot of Rings over Sex
ggplot(abalone) + aes(Rings, fill = Sex) +  geom_density(alpha = 0.7)

```

As we can see Females have higher number of rings and thus we could say that they live longer.

### 2.4 WholeWeight

```{r}
# Summary statistics
summary(abalone$Wholeweight)
```







## 3. Model Selection 


### 3.1 Testing and Evaluation metrics

```{r}
# Algorithms using 10-fold cross validation
control <- trainControl(method="repeatedcv", number=10, repeats=3)
metric <- "RMSE"
seed <- 888
```


### 3.2 Model Generation (non-ensemble)


```{r}
# GLM
set.seed(seed)
fit.glm <- train(Rings~., data=abalone, method="glm", metric=metric, trControl=control)
# LM
set.seed(seed)
fit.lm <- train(Rings~., data=abalone, method="lm", metric=metric, trControl=control)
# SVM
set.seed(seed)
fit.svm <- train(Rings~., data=abalone, method="svmRadial", metric=metric, trControl=control)
# kNN
set.seed(seed)
fit.knn <- train(Rings~., data=abalone, method="knn", metric=metric, trControl=control)

```


### 3.3 Algorithm Comparision


```{r}
results <- resamples(list(SVM=fit.svm, kNN=fit.knn, GLM=fit.glm, LM=fit.lm))
summary(results)
```



```{r}
dotplot(results)
```


# 4. Results Validation

```{r}
index <- best(fit.svm$results, metric, maximize=TRUE)
config <- fit.svm$results[index,]
print(config)
```

